Characterizing edges in signed and vector-valued graphs

### 2018-03-31 edge/model.tex
shorten table captions in edge chapter
typo

\begin{enumerate}[1),nosep,leftmargin=*]
  \item We naturally describe an edge $(u,v)$ through the profiles of its endpoints. More precisely,
    we combine $x_u$ and $x_v$ by an operator $s$ defined as:
    \begin{equation*}
      s(u, v) = x_u \circ x_v = s_{uv} \in \Rbb^d \, ,
    \end{equation*}
    where $\circ$ stands for the Hadamard (or component wise) product. As showed in
    \autoref{tab:edge_sim}, for each dimension $i \in \rangesd$, the vector $s_{uv}$ holds a value
    indicating whether the profiles of $u$ and $v$ agree, disagree or are indifferent along that
    dimension.
    \begin{table}[hb]
      \centering
      % \small
      % \captionsetup{font=small,type=table}
      \caption{\small We show the effect of the Hadamard product in combining information from two
        users $u$ and $v$. Instead of actual numbers, in this table we use $+$ to denote a large
        positive value, $-$ to denote a large negative value, $0$ to denote a small absolute value
        and $\star$ to denote an arbitrary value.
      \label{tab:edge_sim}}
      \begin{tabular}{ccc}
        \toprule
        ${x_u}_{;i}$ & ${x_v}_{;i}$ & ${s_{uv}}_{;i}$ \\
        \midrule
        $+$          & $+$          & $+$             \\
        $-$          & $-$          & $+$             \\
        $+$          & $-$          & $-$             \\
        $0$          & $\star$      & $0$             \\
        \bottomrule
      \end{tabular}
    \end{table}
  \item The $k$ explanations take the form of bounded norm vectors $\mathcal{D}_k = \{w_1, w_2,
    \ldots, w_k\} \subset \dball$, where $\dball$ is the $\ell_2$-unit ball of $\Rbb^d$. We call such a
    vector $w_\ell$, for $\ell \in \rangesk$, a \emph{direction}. As we shall explain more formally
    later on, those directions are exactly the parameters learned while solving our problem.
  \item Given an edge $(u,v)$, all the $k$ directions $\{w_1, w_2, \ldots, w_k\}$ provide a possible
    explanation of why $u$ and $v$ are connected. We score these explanations by assigning them a
    \emph{goodness} of explanation\footnote{Whereas the name is inspired by the concept of
    \emph{goodness of fit}, there is no formal relation.} defined as:
    \begin{equation*}
      g(s_{uv}, w_\ell) = {s_{uv}}^T w_\ell \, .
    \end{equation*}
    As the $w$ notation hints, a direction can be interpreted as a set of weights. These weights
    indicates to which extent each attribute contributes to the connection, according to that
    direction. For a given set of directions $\mathcal{D}_k$, the classification of edges is
    naturally performed by the following operator $\mathcal{E}$, returning the index of the
    direction with the maximal goodness:
    \begin{equation*}
      \begin{array}[t]{lrcl}
        \mathcal{E} : & E \times \left( \Rbb^d \times \ldots \times \Rbb^d \right)
                      & \longrightarrow & \rangesk \\
                      & (u,v), \mathcal{D}_k & \longmapsto
                      & \argmax_{\ell \in \rangesk} g(s_{uv}, w_\ell)
      \end{array}
    \end{equation*}
    Because it will be clear from context, we lighten the notation and simply write $\mathcal{E}(u,
    v)$ instead of $\mathcal{E}\left((u, v), \mathcal{D}_k \right)$.
\end{enumerate}

@@

Second, we also make the assumption that each node $u$ is only involved in $k_\mathrm{local} < k$
different directions. The intuition is that when $d$ is large enough, users have to focus their
energy and can only express their interest in a few number of dimensions. In other words, all the
edges incident to $u$ can only be associated with one of $k_\mathrm{local}$ directions. For
some values of
$k_\mathrm{local}$, this is a \NPc{} graph colouring problem and thus cannot necessarily be enforced
exactly. Still, it can be seen as a learning bias and will be used to generate synthetic data.
Furthermore, we can also express it as an optimization constraint. Specifically, suppose that for
an edge $(u,v)$ associated with the direction $w_i \in \mathcal{D}_k$, we have a $k$-dimensional
vector $y_{uv}$ equal to $e_i$ (that is all zero except for the \ith{} component set to one). Let
us stack all such vectors into a $|E| \times k$ matrix $Y$, and let $C \in \Rbb^{|V| \times |E|}$ be
the incidence matrix of $G$\footnote{The incidence matrix of a graph $G$, usually denoted by $B$, is
the $|V| \times |E|$ matrix such that $B_{i,j} = 1$ if the node $v_i$ and edge $e_j$ are incident
and 0 otherwise.}, with $c_u$ being its $u^{th}$ row. Then $c_u^TY$ is a $k$-dimensional vector that counts
how many times each direction is incident to node $u$. We thus want the $\ell_0$ norm of $c_u^TY$
(\ie the number of its non-zero component) to be upper bounded by $k_\mathrm{local}$.
This is not a convex constraint but we can relax it, for such sparsity inducing problems have been
well studied~\autocite{sparseOptim12}. Namely, we let $y_{uv}$ be a softmax membership
$k$-dimensional vector, that is we set, $\forall i \in \rangesk$,
\begin{equation*}
  {y_{uv}}_{;i} = \frac{\exp(\beta g(s_{uv}, w_i))}{\sum_{j=1}^k \exp(\beta g(s_{uv}, w_j))}\,,
\end{equation*}
with $\beta>0$ a large
positive constant.  And we replace the $\ell_0$ norm by a $\ell_1$ norm (or even a $k$-support
norm~\autocite{KsupportNorm12}), yielding the following $\mathcal{L}_{\mathrm{local}}$ term to be
minimized:
\begin{equation}
  \label{eq:edge_local_loss}
  \mathcal{L}_{\mathrm{local}} =
  \sum_{u \in V} \left\| \sum_{v \in \nei(u)} y_{uv} \right\|_1 =
  \sum_{u \in V} \left( \sum_{i = 1}^k \sum_{v \in \nei(u)}
  \frac{\exp(\beta g(s_{uv}, w_i))}{\sum_{j=1}^k \exp(\beta g(s_{uv}, w_j))} \right)
\end{equation}
It is reasonable to further expect that two neighboring nodes share one common direction among their
$k_\mathrm{local}$ ones. One way to ensure this condition is to imagine that the set of allowed
$k_\mathrm{local}$ directions changes smoothly over the graph. This property depends of the
topology, as we can see by considering two extreme cases: a line and a complete graph. On the line,
once a node has picked two allowed directions from its two neighbors, the remaining ones can be
chosen arbitrarily. On the complete graph on the other hand, this choice is must more constrained.

>>>>

\begin{enumerate}[1),nosep,leftmargin=*]
  \item We naturally describe an edge $(u,v)$ through the profiles of its endpoints. More precisely,
    we combine $x_u$ and $x_v$ by an operator $s$ defined as:
    \begin{equation*}
      s(u, v) = x_u \circ x_v = s_{uv} \in \Rbb^d \, ,
    \end{equation*}
    where $\circ$ stands for the Hadamard (or component wise) product. As showed in
    \autoref{tab:edge_sim}, for each dimension $i \in \rangesd$, the vector $s_{uv}$ holds a value
    indicating whether the profiles of $u$ and $v$ agree, disagree or are indifferent along that
    dimension.
    \begin{table}[hb]
      \centering
      % \small
      % \captionsetup{font=small,type=table}
      \caption[Combining node profiles through the Hadamard product]{%
        \small We show the effect of the Hadamard product in combining information from two
        users $u$ and $v$. Instead of actual numbers, in this table we use $+$ to denote a large
        positive value, $-$ to denote a large negative value, $0$ to denote a small absolute value
        and $\star$ to denote an arbitrary value.
      \label{tab:edge_sim}}
      \begin{tabular}{ccc}
        \toprule
        ${x_u}_{;i}$ & ${x_v}_{;i}$ & ${s_{uv}}_{;i}$ \\
        \midrule
        $+$          & $+$          & $+$             \\
        $-$          & $-$          & $+$             \\
        $+$          & $-$          & $-$             \\
        $0$          & $\star$      & $0$             \\
        \bottomrule
      \end{tabular}
    \end{table}
  \item The $k$ explanations take the form of bounded norm vectors $\mathcal{D}_k = \{w_1, w_2,
    \ldots, w_k\} \subset \dball$, where $\dball$ is the $\ell_2$-unit ball of $\Rbb^d$. We call such a
    vector $w_\ell$, for $\ell \in \rangesk$, a \emph{direction}. As we shall explain more formally
    later on, those directions are exactly the parameters learned while solving our problem.
  \item Given an edge $(u,v)$, all the $k$ directions $\{w_1, w_2, \ldots, w_k\}$ provide a possible
    explanation of why $u$ and $v$ are connected. We score these explanations by assigning them a
    \emph{goodness} of explanation\footnote{Whereas the name is inspired by the concept of
    \emph{goodness of fit}, there is no formal relation.} defined as:
    \begin{equation*}
      g(s_{uv}, w_\ell) = {s_{uv}}^T w_\ell \, .
    \end{equation*}
    As the $w$ notation hints, a direction can be interpreted as a set of weights. These weights
    indicates to which extent each attribute contributes to the connection, according to that
    direction. For a given set of directions $\mathcal{D}_k$, the classification of edges is
    naturally performed by the following operator $\mathcal{E}$, returning the index of the
    direction with the maximal goodness:
    \begin{equation*}
      \begin{array}[t]{lrcl}
        \mathcal{E} : & E \times \left( \Rbb^d \times \ldots \times \Rbb^d \right)
                      & \longrightarrow & \rangesk \\
                      & (u,v), \mathcal{D}_k & \longmapsto
                      & \argmax_{\ell \in \rangesk} g(s_{uv}, w_\ell)
      \end{array}
    \end{equation*}
    Because it will be clear from context, we lighten the notation and simply write $\mathcal{E}(u,
    v)$ instead of $\mathcal{E}\left((u, v), \mathcal{D}_k \right)$.
\end{enumerate}

@@

Second, we also make the assumption that each node $u$ is only involved in $k_\mathrm{local} < k$
different directions. The intuition is that when $d$ is large enough, users have to focus their
energy and can only express their interest in a few number of dimensions. In other words, all the
edges incident to $u$ can only be associated with one of $k_\mathrm{local}$ directions. For
some values of
$k_\mathrm{local}$, this is a \NPc{} graph colouring problem and thus cannot necessarily be enforced
exactly. Still, it can be seen as a learning bias and will be used to generate synthetic data.
Furthermore, we can also express it as an optimization constraint. Specifically, suppose that for
an edge $(u,v)$ associated with the direction $w_i \in \mathcal{D}_k$, we have a $k$-dimensional
vector $y_{uv}$ equal to $e_i$ (that is all zero except for the \ith{} component set to one). Let
us stack all such vectors into a $|E| \times k$ matrix $Y$, and let $C \in \Rbb^{|V| \times |E|}$ be
the incidence matrix of $G$\footnote{The incidence matrix of a graph $G$, usually denoted by $B$, is
the $|V| \times |E|$ matrix such that $B_{i,j} = 1$ if the node $v_i$ and edge $e_j$ are incident
and 0 otherwise.}, with $c_u$ being its $u^{th}$ row. Then $c_u^TY$ is a $k$-dimensional vector that counts
how many times each direction is incident to node $u$. We thus want the $\ell_0$ norm of $c_u^TY$
(\ie the number of its non-zero component) to be upper bounded by $k_\mathrm{local}$.
This is not a convex constraint but we can relax it, for such sparsity inducing problems have been
well studied~\autocite{sparseOptim12}. Namely, we let $y_{uv}$ be a softmax membership
$k$-dimensional vector, that is we set, $\forall i \in \rangesk$,
\begin{equation*}
  {y_{uv}}_{;i} = \frac{\exp(\beta g(s_{uv}, w_i))}{\sum_{j=1}^k \exp(\beta g(s_{uv}, w_j))}\,,
\end{equation*}
with $\beta>0$ a large
positive constant.  And we replace the $\ell_0$ norm by a $\ell_1$ norm (or even a $k$-support
norm~\autocite{KsupportNorm12}), yielding the following $\mathcal{L}_{\mathrm{local}}$ term to be
minimized:
\begin{equation}
  \label{eq:edge_local_loss}
  \mathcal{L}_{\mathrm{local}} =
  \sum_{u \in V} \left\| \sum_{v \in \nei(u)} y_{uv} \right\|_1 =
  \sum_{u \in V} \left( \sum_{i = 1}^k \sum_{v \in \nei(u)}
  \frac{\exp(\beta g(s_{uv}, w_i))}{\sum_{j=1}^k \exp(\beta g(s_{uv}, w_j))} \right)
\end{equation}
It is reasonable to further expect that two neighboring nodes share one common direction among their
$k_\mathrm{local}$ ones. One way to ensure this condition is to imagine that the set of allowed
$k_\mathrm{local}$ directions changes smoothly over the graph. This property depends of the
topology, as we can see by considering two extreme cases: a line and a complete graph. On the line,
once a node has picked two allowed directions from its two neighbors, the remaining ones can be
chosen arbitrarily. On the complete graph on the other hand, this choice is much more constrained.
